---
title: "Getting started with lenguar"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with lenguar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lenguar)
```

lenguar wraps [lengua](https://github.com/jprybylski/lengua) — a git-backed library for
templated text and snippets — via Rust FFI (extendr), so every function here runs the same
`lengua-core` engine as the standalone `lengua` CLI, just without a subprocess in between.

## A store is a git repo

Every lenguar function takes a `path` to a *store*: a directory containing a `templates/`
folder and its own `.git` history. Create one with `lq_init()`:

```{r}
store <- withr::local_tempdir()
lq_init(store)
```

## Adding templates

`lq_add()` writes a template body plus optional frontmatter (`title`, and any other
`fields`) and commits the change:

```{r}
lq_add(
  store, "letters/thank-you.md",
  "Dear {{ name }},\n\nThank you for {{ reason }}.",
  title = "Thank You",
  fields = c(tone = "formal")
)
```

Calling `lq_add()` again for the same `name` records a new version — lenguar doesn't have a
separate versioning mechanism, it's just git commits (see [History](#history) below).

## Rendering

`lq_get()` substitutes `{{ variable }}` placeholders
([minijinja](https://github.com/mitsuhiko/minijinja)/Jinja2 syntax) using the named character
vector passed as `vars`:

```{r}
lq_get(store, "letters/thank-you.md", vars = c(name = "Ada", reason = "your thoughtful review"))
```

Pass `raw = TRUE` to get the unrendered body back instead (useful for round-tripping a template
into another store, or into an editor).

## Listing and searching

`lq_list()` returns every template as a data frame; `lq_search()` filters by frontmatter field
(every `fields` pair must match — an AND, not an OR):

```{r}
lq_add(store, "letters/casual-note.md", "Hey {{ name }}!", fields = c(tone = "casual"))

lq_list(store)

lq_search(store, c(tone = "formal"))
```

Both return a `lenguar_templates` data frame, so anything that works on a data frame (subsetting,
`dplyr::filter()`, ...) works here too.

## History {#history}

Because every `lq_add()` is a git commit, `lq_log()` and `lq_diff()` give you real history for
free:

```{r}
lq_add(store, "letters/thank-you.md",
  "Dear {{ name }},\n\nThank you so much for {{ reason }}.",
  title = "Thank You", fields = c(tone = "formal")
)

lq_log(store, "letters/thank-you.md")

lq_diff(store, "letters/thank-you.md")
```

## Tagging revisions {#tagging}

`lq_tag()` points a named tag at a specific revision of a template — the current one, or (via
`rev`) a prior one, which is how you *retroactively* tag a revision after the fact:

```{r}
lq_add(store, "letter.md", "We had so much fun.", message = "past tense")
lq_add(store, "letter.md", "We will have so much fun.", message = "future tense")

lq_tag(store, "letter.md", "tense-future")
lq_tag(store, "letter.md", "tense-past", rev = "HEAD~1")

lq_tag_list(store, "letter.md")
```

These are lenguar's own tags (`refs/lengua/tags/<template>/<tag>`), not git tags — they're
scoped per template, so the same tag name can independently exist on several templates. A tag
name works anywhere `lq_get()`'s `rev` or `lq_diff()`'s `from`/`to` accept a revision:

```{r}
lq_get(store, "letter.md", rev = "tense-past")

lq_diff(store, "letter.md", from = "tense-past", to = "tense-future")
```

## Consuming an existing library

Most of the time you're not building a template library from scratch — you're using one a team
already maintains. `lq_init()`'s `from_dir`/`from_repo` arguments adopt an existing store
instead of starting empty, mirroring the `lengua` CLI's `init --from-dir`/`--from-repo`.
`from_dir` (and `from_repo` when it names a local path) copies the source on the filesystem,
preserving its full history and tags:

```{r}
copy <- withr::local_tempdir()
lq_init(copy, from_dir = store)

lq_log(copy, "letter.md")
lq_tag_list(copy, "letter.md")
```

`from_repo` against a real remote host clones over the network instead:

```r
lq_init(store, from_repo = "acme-org/team-templates")
lq_init(store, from_repo = "acme-org/team-templates/letters@v2")
```

See [lengua's own docs on consuming an existing
library](https://jprybylski.github.io/lengua/consuming.html) for the full walkthrough, including
GitHub Enterprise hosts and importing a subdirectory.

## Error handling

Every lenguar function that talks to the underlying store raises a classed condition on
failure — `lenguar_store_error` for problems from lengua-core (a missing template, an invalid
revision, ...) and `lenguar_input_error` for bad arguments caught on the R side (an unnamed
`fields`/`vars` vector, an empty `search()` filter). Both inherit from `lenguar_error`, so you
can catch either specifically or both at once:

```{r, error = TRUE}
tryCatch(
  lq_get(store, "does-not-exist.md"),
  lenguar_store_error = function(e) cat("store error:", conditionMessage(e), "\n")
)
```

## Where to next

See `?lq_add`, `?lq_get`, `?lq_list`, `?lq_search`, `?lq_log`, `?lq_diff`, `?lq_tag`,
`?lq_tag_list`, `?lq_tag_rm`, and `?lq_init` for full per-function reference, and [lengua's own
docs](https://jprybylski.github.io/lengua/) for the template syntax, frontmatter model, and
storage design that lenguar is a thin FFI layer over.
